---
title: "Assessment 3 of Mihiravi"
author: "Mihiravi"
date: "21/05/2021"
output: html_document
---

##Part 1

#Question 1

#Download the file “gene_expression2.tsv” which contains RNA-seq count data for two samples of interest. Read in the file, making the gene accession numbers the row names. Show a table of values for the first six genes

```{r gene_expression.tsv}

download.file("https://raw.githubusercontent.com/markziemann/SLE712_files/master/assessment_task3/bioinfo_asst3_part1_files/gene_expression.tsv",
              destfile = "gene_expression.tsv")
gene_expression <- read.table("gene_expression.tsv",
                              header = TRUE, 
                              row.names = 1)
str(gene_expression)
head(gene_expression)
# get rows 1-6 of all columns to appear 
gene_expression[1:6,]

```

#Question 2

#new column for analysing row mean values

```{r new column for row means}

# to find the row means
head(gene_expression)
row_means <- rowMeans(gene_expression)
head(row_means)
gene_expression$mean <- c(row_means)
# to show only 6 rows in the table
gene_expression[1:6,]

```

#Question 3

#10 genes with the highest mean expression

```{r highest mean expression}

#bring the highest values to the top of data set by using reverse sorting
reverse_sorted<- gene_expression[order(-gene_expression$mean),]
#to get the 10 highest mean values
reverse_sorted[1:10, "mean"]

```


#Question 4

#number of genes with a mean <10
```{r mean <10 number of genes}
number_genes <- sum(gene_expression$mean <10)
number_genes
```


#Question 5

#histogram plot of the mean values in png format

```{r histogram}
hist (gene_expression$mean, 
     main = "Histogram plot of the mean values of genes",
     xlab = "Mean values",
     col = "blue")

# to save as a png format 
png(file="Histogram plot of the mean values of genes")
```

#Question 6

#downloading the growth_data.csv file

```{r growth_data.csv file}

download.file("https://raw.githubusercontent.com/markziemann/SLE712_files/master/assessment_task3/bioinfo_asst3_part1_files/growth_data.csv", destfile = "growth_data.csv")
Growth_data <- read.csv("growth_data.csv", header = TRUE, stringsAsFactors = FALSE)
head(Growth_data)
colnames(Growth_data)

```


#Question 7

#mean and standard deviation of tree circumference at the start and end of the study at both sites

```{r mean and SD  of 2004 and 2019}

# for the northeast data set for 2004
NE <- subset(Growth_data, Site== "northeast")
head(NE)
str(NE)
# to find the mean of 2004 for NE
mean_2004_NE <- mean(NE$Circumf_2004_cm)
mean_2004_NE
# to find the SD of 2004 for NE
SD_2004_NE <- sd(NE$Circumf_2004_cm)
SD_2004_NE

## for the southwest data set for 2004
SW <- subset(Growth_data, Site=="southwest")
head(SW)
str(SW)
# to find the mean of 2004 for SW
mean_2004_SW <- mean(SW$Circumf_2004_cm)
mean_2004_SW
# to find the SD of 2004 for SW
SD_2004_SW <- sd(SW$Circumf_2004_cm)
SD_2004_SW

##For year 2019

# for the northeast data set 
# to find the mean of 2004 for NE
mean_2019_NE <- mean(NE$Circumf_2019_cm)
mean_2019_NE
# to find the SD of 2004 for NE
SD_2019_NE <- sd(NE$Circumf_2019_cm)
SD_2019_NE

# # for the southwest data set for 2019
# to find the mean of 2019 for SW
mean_2019_SW <- mean(SW$Circumf_2019_cm)
mean_2019_SW
#to find the SD of 2019 for SW
SD_2019_SW <- sd(SW$Circumf_2019_cm)
SD_2019_SW

```

#Question 8

#box plot of tree circumference at the start and end of the study at both sites

```{r box plot of tree circumference 2004,2019}

boxplot(NE$Circumf_2004_cm, SW$Circumf_2004_cm, NE$Circumf_2019_cm, SW$Circumf_2019_cm,
    names = c("North east-2004","South west-2004","North east-2019","South west-2019"),
    ylab=("Circumference (cm)"),
    xlab= ("Year"),
    col = "blue",
    main= "Tree circumference at the start and end of the study")

```

#Question 9

#mean growth over the past 10 years at each site

```{r mean growth over past 10 years}
#for northeast site
growth_2009_NE <- mean(NE$Circumf_2009_cm)
head(growth_2009_NE)
growth_2019_NE <- mean(NE$Circumf_2019_cm)
head(growth_2019_NE)
#mean growth in northeast site
treegrowth_NE <- (growth_2019_NE -growth_2009_NE )
treegrowth_NE

#for southeast site

growth_2019_SW <- mean(SW$Circumf_2019_cm)
head(growth_2019_SW)
growth_2009_SW <- mean(SW$Circumf_2009_cm)
head(growth_2009_SW)
#mean growth in southeast site
treegrowth_SW <- (growth_2019_SW-growth_2009_SW)
treegrowth_SW

```

#Question 10 

#t.test and wilcox.test functions to estimate the p-value that the 10 year growth is di

```{r t.test and wilcox.test}
# ten year growth in Northeast
tenyear_treegrowth_NE <- (NE$Circumf_2019_cm- NE$Circumf_2009_cm)
tenyear_treegrowth_NE

# ten year growth in Southwest
tenyear_treegrowth_SW <-(SW$Circumf_2019_cm -SW$Circumf_2009_cm)
tenyear_treegrowth_SW

# t.test
t.test(tenyear_treegrowth_NE,tenyear_treegrowth_SW )
# wilcox.test
wilcox.test (tenyear_treegrowth_NE,tenyear_treegrowth_SW)

```

##### Part 2


