---
title: "Assessment 3- Daisy Ta"
author: "Daisy Ta"
date: "5/17/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---
#PART 1
#Question 1: Here is the downloaded file "gene_expression" which contains RNA-seq count data for two samples of interest...... And finally, we used "str" to check that the structure is what they wanted, and then we output, the first six lines six rows using the head command.
```{r gene_expression.tsv}
download.file("https://raw.githubusercontent.com/markziemann/SLE712_files/master/assessment_task3/bioinfo_asst3_part1_files/gene_expression.tsv",
              destfile= "gene_expression.tsv")
gene_expression <- read.table("gene_expression.tsv", header=TRUE, row.names=1)
str(gene_expression)
head(gene_expression)
# a table of values for the frst six genes.
gene_expression[1:6,]
```

#Question 2:
```{r gene_expression.tsv}
head(gene_expression)
Col_mean <- rowMeans(gene_expression)
Col_mean
gene_expression$mean <- c(Col_mean)
gene_expression
gene_expression[1:6,]


```
#Question 3: List the 10 genes with the highest mean expression.
```{r gene_expression.tsv}
order(gene_expression$mean)
sorted_gene_expression <- gene_expression[order(gene_expression$mean), ]
top10genes <- row.names(tail(sorted_gene_expression, 10))
top10genes


```
## Question 4: Determine the number of genes with a mean <10.

```{r gene_expression.tsv}
number_genes <- sum(gene_expression$mean < 10)
cat("The number of genes with a mean < 10:", number_genes)

```

#Question 5: Make a histogram plot of the mean values in png format and paste it into your report.

```{r gene_expression.tsv}
png(file="histogram.png")
hist(gene_expression$mean, main= "Histogram plot of the mean values", xlab= "Mean values", col="green")



```
#Question 6: Import this csv file into an R object. What are the column names?
```{r growth_data.csv}
download.file("https://raw.githubusercontent.com/markziemann/SLE712_files/master/assessment_task3/bioinfo_asst3_part1_files/growth_data.csv",
              destfile= "growth_data.csv")
growth_data <- read.csv("growth_data.csv", header=TRUE,stringsAsFactors = FALSE)
head(growth_data)
colnames(growth_data)
```

#Question 7: Calculate the mean and standard deviation of tree circumference at the start and end of the study at both sites.
```{r growth_data.csv}
#The mean and standard deviation of tree circumference at the start (in 2004)
mean_2004 <- mean(growth_data$Circumf_2004_cm)
mean_2004
cat("Mean in 2004:", mean_2004)
sd_2004 <- sd(growth_data$Circumf_2004_cm)
sd_2004
cat("SD in 2004:", sd_2004)

#The mean and standard deviation of tree circumference at the end (in 2019)
mean_2019 <- mean(growth_data$Circumf_2019_cm)
mean_2019
cat("Mean in 2019:", mean_2019)
sd_2019 <- sd(growth_data$Circumf_2019_cm)
sd_2019
cat("SD in 2019:", sd_2019)

```
#Question 8: Make a box plot of tree circumference at the start and end of the study at both sites.
```{r growth_data.csv}
x_vals <- growth_data$Circumf_2004_cm
y_vals <- growth_data$Circumf_2019_cm
boxplot(x_vals, y_vals,
main = "Box plot of tree circumference",
names = c("2004", "2019"),
xlab = "Year",
ylab = "Tree circumference (cm)",
col = "orange"
)
```

#Question 9: Calculate the mean growth over the past 10 years at each site.
```{r growth_data.csv}
Present <- mean(growth_data$Circumf_2014_cm)
Present
Past <- mean(growth_data$Circumf_2004_cm)
Past
growth_data_growth <- Present - Past
growth_data_growth
head(growth_data_growth)
str(growth_data)


```