---
title: "Assessment 3- Daisy Ta"
author: "Daisy Ta"
date: "5/17/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---
#PART 1
#Question 1: Here is the downloaded file "gene_expression" which contains RNA-seq count data for two samples of interest...... And finally, we used "str" to check that the structure is what they wanted, and then we output, the first six lines six rows using the head command.
```{r gene_expression.tsv}
download.file("https://raw.githubusercontent.com/markziemann/SLE712_files/master/assessment_task3/bioinfo_asst3_part1_files/gene_expression.tsv",
              destfile= "gene_expression.tsv")
gene_expression <- read.table("gene_expression.tsv", header=TRUE, row.names=1)
str(gene_expression)
head(gene_expression)
# a table of values for the frst six genes.
gene_expression[1:6,]
```

#Question 2:
```{r gene_expression.tsv}
head(gene_expression)
Col_mean <- rowMeans(gene_expression)
Col_mean
gene_expression$mean <- c(Col_mean)
gene_expression
gene_expression[1:6,]


```
#Question 3: List the 10 genes with the highest mean expression.
```{r gene_expression.tsv}
order(gene_expression$mean)
sorted_gene_expression <- gene_expression[order(gene_expression$mean), ]
top10genes <- row.names(tail(sorted_gene_expression, 10))
top10genes


```
## Question 4: Determine the number of genes with a mean <10.

```{r gene_expression.tsv}
number_genes <- sum(gene_expression$mean < 10)
cat("The number of genes with a mean < 10:", number_genes)

```

#Question 5: Make a histogram plot of the mean values in png format and paste it into your report.

```{r gene_expression.tsv}
png(file="histogram.png")
hist(gene_expression$mean, main= "Histogram plot of the mean values", xlab= "Mean values", col="green")



```
#Question 6: Import this csv file into an R object. What are the column names?
```{r growth_data.csv}
download.file("https://raw.githubusercontent.com/markziemann/SLE712_files/master/assessment_task3/bioinfo_asst3_part1_files/growth_data.csv",
              destfile= "growth_data.csv")
growth_data <- read.csv("growth_data.csv", header=TRUE,stringsAsFactors = FALSE)
head(growth_data)
colnames(growth_data)
```

#Question 7: Calculate the mean and standard deviation of tree circumference at the start and end of the study at both sites.
```{r growth_data.csv}
#The mean and standard deviation of tree circumference at the start (in 2004)
mean_2004 <- mean(growth_data$Circumf_2004_cm)
mean_2004
cat("Mean in 2004:", mean_2004)
sd_2004 <- sd(growth_data$Circumf_2004_cm)
sd_2004
cat("SD in 2004:", sd_2004)

#The mean and standard deviation of tree circumference at the end (in 2019)
mean_2019 <- mean(growth_data$Circumf_2019_cm)
mean_2019
cat("Mean in 2019:", mean_2019)
sd_2019 <- sd(growth_data$Circumf_2019_cm)
sd_2019
cat("SD in 2019:", sd_2019)

```
#Question 8: Make a box plot of tree circumference at the start and end of the study at both sites.
```{r growth_data.csv}
x_vals <- growth_data$Circumf_2004_cm
y_vals <- growth_data$Circumf_2019_cm
boxplot(x_vals, y_vals,
main = "Box plot of tree circumference",
names = c("2004", "2019"),
xlab = "Year",
ylab = "Tree circumference (cm)",
col = "orange"
)
```

#Question 9: Calculate the mean growth over the past 10 years at each site.
```{r growth_data.csv}
# Get the tree circumference values of each site
northeast <- subset(growth_data, Site=="northeast")
southwest <- subset(growth_data, Site=="southwest")
# Calculate the mean values of each site over past 10 years
na_2019 <- mean(northeast$Circumf_2019_cm)
sw_2019 <- mean(southwest$Circumf_2019_cm)
na_2009 <- mean(northeast$Circumf_2009_cm)
sw_2009 <- mean(southwest$Circumf_2009_cm)
# Calculate the mean growth of each site
na_mean_growth <- na_2019 - na_2009
sw_mean_growth <- sw_2019 - sw_2009
cat("The mean growth over the past 10 years of Northeast:", na_mean_growth)
cat("The mean growth over the past 10 years of Southwest:", sw_mean_growth)
```

#Question 10: Use the t.test and wilcox.test functions to estimate the p-value that the 10 year growth is different at the two sites.
```{r growth_data}
#get the growth values of each site
na_growth <- northeast$Circumf_2019_cm - northeast$Circumf_2009_cm
sw_growth <- southwest$Circumf_2019_cm - southwest$Circumf_2009_cm
#run t.test
t <- t.test(na_growth, sw_growth)
t
#get p-value of t.test
t_pvalue <- t$p.value
cat("p-value of t.test:", t_pvalue)
#run wilcox.test
wilcox <- wilcox.test(na_growth, sw_growth)
wilcox
# get p-value of wilcox.test
wilcox_pvalue <- wilcox$p.value
cat("p-value of wilcox.test:", wilcox_pvalue)
```
